---
title: "BIOS707 | Problem Set 01"
author: "Kuei-Yueh (Clint) Ko"
output:
  html_notebook: default
  pdf_document: default
---

# Set environment

This report aim to recreate the table one and table two in the study *"Operation Timing and 30-Dayrtality After Elective General Surgery"*. The packages required in this report include **dplyr**, **tableone**, **labelled**.

```{r, message=FALSE}
### load required packages
library(dplyr)
library(tableone)
library(labelled)
```

# Import data

The study data are available in a number of formats on the website. Load the RData object directly from the website.

```{r}
### load R dataset
load(url("https://www.causeweb.org/tshs/datasets/surgery_timing.Rdata"))
head(stata_data)
```


# Table 01

Recreate table 1 of the study. The table 1 is **the summary of baseline risk factors for 32,001 general surgical patients**.

```{r}
## intialization
df <- stata_data

## Make categorical variables factors
varsToFactor <- c("gender", "race", "asa_status", 
          "baseline_cancer",    # Cancer
          "baseline_cvd",       # Cardiovascular/cerebrovascular disease 
          "baseline_dementia",  # Dementia
          "baseline_diabetes",  # Diabetes
          "baseline_digestive", # Digestive disease
          "baseline_osteoart",  # Osteoarthritis
          "baseline_psych",     # Psychiatric disorder
          "baseline_pulmonary") # Pulmonary disease
          
    
df[varsToFactor] <- lapply(df[varsToFactor], factor)

## set levels
df$asa_status <- factor(df$asa_status, levels = c("I-II", "III", "IV-VI", ""))

## set label of variables
var_label(df$age)                <- "Age"
var_label(df$gender)             <- "Gender"
var_label(df$race)               <- "Race"
var_label(df$asa_status)         <- "ASA physical status"
var_label(df$bmi)                <- "Body mass index (kg/m^2)"
var_label(df$baseline_cancer)    <-  "Cancer"
var_label(df$baseline_cvd)       <- "Cardiovascular/cerebrovascular disease"
var_label(df$baseline_dementia)  <- "Dementia"
var_label(df$baseline_diabetes)  <- "Diabetes"
var_label(df$baseline_digestive) <- "Digestive disease"
var_label(df$baseline_osteoart)  <- "Osteoarthritis"
var_label(df$baseline_psych)     <- "Psychiatric disorder"
var_label(df$baseline_pulmonary) <- "Pulmonary disease"
var_label(df$baseline_charlson)  <- "Charison Comorbidity Index"
var_label(df$mortality_rsi)      <- "Risk Stratification Index (30-day mortality)"
var_label(df$complication_rsi)   <- "Risk Stratification Index (in-hospital complications)"

## Create a variable list
vars <- c("age",                # Age
          "gender",             # Gender
          "race",               # Race
          "asa_status",         # ASA physical status
          "bmi",                # Body mass index (kg/m^2)
          "baseline_cancer",    # Cancer
          "baseline_cvd",       # Cardiovascular/cerebrovascular disease 
          "baseline_dementia",  # Dementia
          "baseline_diabetes",  # Diabetes
          "baseline_digestive", # Digestive disease
          "baseline_osteoart",  # Osteoarthritis
          "baseline_psych",     # Psychiatric disorder
          "baseline_pulmonary", # Pulmonary disease
          "baseline_charlson",  # Charison Comorbidity Index
          "mortality_rsi",      # Risk Stratification Index (30-day mortality)
          "complication_rsi")   # Risk Stratification Index (in-hospital complications)

## nonnormal continuous variables
vars_nonnorm <- c("bmi", "baseline_charlson", "mortality_rsi", "complication_rsi")

## Create Table 1 stratified by trt
tableOne <- CreateTableOne(data = df, vars = vars)

## Format
print(tableOne, 
      contDigits = 1,
      nonnormal  = vars_nonnorm,
      varLabels  = TRUE,
      explain    = FALSE,
      dropEqual  = TRUE)
```

# Table 02

Recreate table two of the study. The table 2 is **frequencies of case start times by hour of day, day of week, month of Year, and moon phase**.

```{r}
### initialization
df <- stata_data

### preprocess and set factor levels
df$hour <- floor(df$hour) %>% 
    as.character %>% 
    sapply(., function(x){return(ifelse(x == "19", "18", x))}) %>% 
    paste0(., ":00") %>%
    factor(., levels = paste0(6:18, ":00"))

# set week abbreviation and full name
week_abb = c("Mon", "Tue", "Wed", "Thu", "Fri")
week_name = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")

# Day of week
df$dow <- df$dow %>% 
    match(., week_abb) %>%          # get the index of each element by matching abbr.
    `[`(week_name, .) %>%           # change the element into full name
    factor(., levels = week_name)   # specify factor level

# Month of year
df$month <- df$month %>%            
    match(., month.abb) %>%         # get the index of each month by matching abbr.
    `[`(month.name, .) %>%          # change the lement into full name
    factor(., levels = month.name)  # specify factor level

# moon phase
df$moonphase <- factor(
    df$moonphase,                   # set as a factor
    levels = c(                     # specify factor level
        "New Moon", 
        "First Quarter", 
        "Full Moon", 
        "Last Quarter"))

### set label of variables
var_label(df$hour)      <- "Factor operation hour"
var_label(df$dow)       <- "Day of week"
var_label(df$month)     <- "Factor month"
var_label(df$moonphase) <- "Phase of moon"

### Create a variable list
vars <- c("hour", "dow", "month", "moonphase") #, "mort30", "complication")

### Create table two
tableTwo <- CreateTableOne(data = df, vars = vars)

## Format
print(tableTwo, 
      contDigits = 1,
      varLabels  = TRUE,
      explain    = FALSE,
      dropEqual  = TRUE)
```

# What does adjusted probability mean in figure 3 and figure 4

**Question**   
Figures 3 and 4 report adjusted probability estimates. What were these probabilities adjusted for? Is this study reproducible based on the published article

**My Answer**    
Based on the statistical analysis in the method part, the thirty-day mortality was adjusted based on Risk Stratification Index (RSI). However, the authors did not mention clearly how and why the probability was adjusted. Instead, the authors only cited the reference of the adjusted method. Since the description is not enough for reader to replicate easily, I believe the calculation of adjusted probability estimates in the study is not reproducible.




