---
title: "BIOS707 | Problem set3"
author: Kuei-Yueh (Clint) Ko
output: 
  html_notebook:
    theme: united
    toc: yes
---

# About the notebook

This the problem set 3 of the course BIOS707. Below are the libraries required for this report.

```{r}
### tools
library(tidyverse)
library(cluster)
library(foreach)
library(clue)
library(pROC)

### plot and display
library(RColorBrewer)
library(pheatmap)
library(grid)
library(gridExtra)
```

# Part I. Simulation

## Q1. Explore Clustering
**Simulate a dataset where n = 100 and p = 50. Used your ICC function from problem set 2 to create 3 (k) distinct clusters in dataset.**

my ICC function to simulate the data
```{r}
simulate_icc <- function(icc, grp_size = 10, n_grp = 5, mu_tot = 0, sig2_a = 1){
    # function to simuate the data from specific ICC
    # =================================
    
    ### calculate sig2_e from the ICC and sig2_a
    sig2_e <- sig2_a * (1 / icc - 1)
    
    ### simulate alpha by sig2_a
    alpha  <- rnorm(n_grp, mean = 0, sd = sig2_a^0.5)
    
    ### simulate error for each group by sig2_e
    df <- sapply(mu_tot + alpha, function(x){
        eps <- rnorm(grp_size, mean = 0, sd = sig2_e^0.5)
        return(x + eps)
    })

    ### set column names
    df <- data.frame(df)
    colnames(df) <- paste("Grp", 1:n_grp)

    ### gather and return
    df <- df %>% gather(Group, Value)
    return(df)
} # end func
```

Demonstrate the results of my function by simulate one paramters only (p = 1)
```{r, fig.width = 10, fig.height = 4}
###
K          <- 3
N          <- 100
ICC1       <- 0.9
SIG2_ALPHA <- 100

###
df <- simulate_icc(ICC1, n_grp = K, grp_size = floor(N / K), sig2_a = SIG2_ALPHA)
df1$ICC <- paste("ICC", "=", ICC1)

df$Group <- factor(df$Group, levels = paste("Grp", 1:K))

###
gp <- ggplot(df, aes(x = Group, y = Value, color = Group)) + 
    geom_jitter(width = 0.2) +
    ggtitle("Simulating random effects with different ICC") +
    theme(
        axis.title.y = element_text(size = 15),
        axis.title.x = element_text(size = 15),
        axis.text.x  = element_text(size = 12, angle = 90, vjust = 0.5),
        strip.text.x = element_text(size = 20, color = "grey30"))
    
print(gp)
```

### Q1 (a) 
**Set your ICC to 0.3 and simulate a dataset. Show your simulation code. Apply K-means clustering and evaluate how well your clusters uncover the true group**


simulate 50 variables (p = 50) with ICC = 0.3
```{r}
### instialization
set.seed(0)
K          <- 3
N          <- 100
ICC1       <- 0.3
SIG2_ALPHA <- 10
P          <- 50

### simulation
dat = replicate(
    P, 
    simulate_icc(
        ICC1, 
        n_grp = K, 
        grp_size = floor(N / K), 
        sig2_a = SIG2_ALPHA),
    simplify = FALSE)

### set binary merge function
my_combine <- function(x, y){
    y = y %>% select(-Group)
    return(bind_cols(x, y))
} # end func

### reduced simulated data into one dataframe
dat = foreach(
    idx = 1:length(dat), 
    .combine = my_combine) %do% {
        df = dat[[idx]]
        df = df %>% dplyr::rename_(
            .dots=setNames(
                names(.), 
                c("Group", paste0("V", idx))))
    }

dat_sim = dat
```

```{r}
head(dat_sim)
```

```{r}
set.seed(0)

unmatch = c()

for (idx in 1:100){
    ###
    mat = dat_sim %>% select(-Group) %>% as.matrix
    
    ###
    clust = kmeans(mat, centers = 3)
    
    ### confusion matrix
    mat_conf = table(clust$cluster, dat_sim$Group)
    
    idx = solve_LSAP(mat_conf, maximum = TRUE)
    num = sum(mat_conf[cbind(seq_along(idx), idx)])
    unmatch = c(unmatch, nrow(mat) - num)
} # end for loop

data.frame(x = 1:length(unmatch), y = unmatch) %>%
    ggplot(., aes(x = x, y = y)) +
    geom_line() +
    geom_point() +
    labs(x = "iteration", y = "number of points mismatch")
```

### Q1 (b)

```{r}
### instialization
K          <- 3
N          <- 100
ICC        <- 0.3
SIG2_ALPHA <- 10
P          <- 50

### set binary merge function
my_combine <- function(x, y){
    y = y %>% select(-Group)
    return(bind_cols(x, y))
} # end func


simulate_data <- function(icc, k, n, sig2_a, p, var_name = "V"){
    ### simulation
    dat = replicate(
        p, 
        simulate_icc(
            icc, 
            n_grp = k, 
            grp_size = floor(n / k), 
            sig2_a = sig2_a),
        simplify = FALSE)
    
    ### reduced simulated data into one dataframe
    dat = foreach(
         idx = 1:length(dat), 
         .combine = my_combine) %do% {
             df = dat[[idx]]
             df = df %>% dplyr::rename_(
                 .dots=setNames(
                names(.), 
                c("Group", paste0(var_name, idx))))
    } # end do in foreach
} # end func

simulate_kmean <- function(dat, iteration){
    unmatch = c()

    for (idx in 1:iteration){
        ###
        mat = dat %>% select(-Group) %>% as.matrix
    
        ###
        clust = kmeans(mat, centers = 3)
    
        ### confusion matrix
        mat_conf = table(clust$cluster, dat_sim$Group)
    
        idx = solve_LSAP(mat_conf, maximum = TRUE)
        num = sum(mat_conf[cbind(seq_along(idx), idx)])
        unmatch = c(unmatch, nrow(mat) - num) 
    } # end for loop
    return(unmatch)
} # end func

simulate_unmatch <- function(icc, k, n, sig2_a, p, iter_km){
    #
    dat_sim = simulate_data(icc, k, n, sig2_a, p)
    unmatch = simulate_kmean(dat_sim, iter_km)
    
    unmatch_events    = sum(unmatch > 0)
    unmatch_points_mu = mean(unmatch)
    unmatch_points_sd = sd(unmatch)
    
    res = c(unmatch_events, unmatch_points_mu, unmatch_points_sd)
    names(res) = c("events", "points_mu", "points_sd")
    
    return(res)
}
```



```{r}
set.seed(0)

iccs = seq(0.1, 1.0, length.out = 10)
ITER_KM = 1000

res = sapply(iccs, function(icc) {
    print(icc)
    res = simulate_unmatch(icc, K, N, SIG2_ALPHA, P, ITER_KM)
    return(res)
}) # end sapply


#for (icc in iccs){
#    print(icc)
#    dat_sim = simulate_data(icc, K, N, SIG2_ALPHA, P)
#    unmatch = simulate_kmean(dat_sim, iter_km)
#    unmatch_events = c(unmatch_events, sum(unmatch > 0))
#    unmatch_points_mu = c(unmatch_points, mean(unmatch))
#    unmatch_points_sd = c(unmatch_points, sd(unmatch))
#} # end for loop
```

```{r}
head(res)
```


```{r, fig.height = 5, fig.width = 5}
dat = data.frame(x = iccs, y = res["events",])
gp1 = ggplot(dat, aes(x = x, y = y)) +
    geom_line() +
    geom_point() +
    labs(x = "ICC", 
         y = "number of time kmeans \nfailed to cluster correctly",
         title = paste("Number of kmeans for each ICC", iter_km))

dat = data.frame(x = iccs, y = res["points_mu",])
gp2 = ggplot(dat, aes(x = x, y = y)) +
    geom_line() +
    geom_point() +
    labs(x = "ICC", 
         y = "Expected points uncover \ninto wrong groups")

grid.arrange(gp1, gp2, nrow = 2)
```


### Q1 (c)

```{r}
simulate_data_noise <- function(icc, q, k, n, sig2_a, p){
    n1 = q
    n2 = p-q
    
    dat1 = simulate_data(icc,    k, n, sig2_a, n1, var_name = "V")
    dat2 = simulate_data(0.0001, k, n, sig2_a, n2, var_name = "N")
    
    tmp = dat2 %>% select(-Group)
    colnames(tmp)
    dat = bind_cols(dat1, tmp)
    
    return(dat)
} # end func

simulate_unmatch_noise <- function(icc, q, k, n, sig2_a, p, iter_km){
    #
    dat_sim = simulate_data_noise(icc, q, k, n, sig2_a, p)
    unmatch = simulate_kmean(dat_sim, iter_km)
    
    unmatch_events    = sum(unmatch > 0)
    unmatch_points_mu = mean(unmatch)
    unmatch_points_sd = sd(unmatch)
    
    res = c(unmatch_events, unmatch_points_mu, unmatch_points_sd)
    names(res) = c("events", "points_mu", "points_sd")
    
    return(unmatch)
}
```

```{r}
simulate_data_noise(0.9, 15, K, N, SIG2_ALPHA, P)
```


```{r}
set.seed(0)
simulate_unmatch_noise(1.0, 49, K, N, SIG2_ALPHA, P, ITER_KM)
```


```{r}
### instialization
K          <- 3
N          <- 100
ICC        <- 0.3
SIG2_ALPHA <- 10
P          <- 50
ITER_KM    <- 500
set.seed(0)

###
iccs = seq(0.1, 1.0, length.out = 10)
qs   = c(1, 5, 10, 20, 25, 49)

res_icc_q = NULL
for (icc in iccs) {
    cat("\nicc", icc)
    for (q in qs) {
        cat("#")
        res  = simulate_unmatch_noise(icc, q, K, N, SIG2_ALPHA, P, ITER_KM)
        res2 = c(res, q, icc)
        names(res2)[4:5] = c("q", "icc")
        
        res_icc_q = rbind(res_icc_q, res2)
        #print(res2)
    } # end inner loop
} # end outer loop
```

```{r}
ggplot(as.data.frame(res_icc_q), 
       aes(x = as.character(q), 
           y = as.character(icc), 
           fill = events)) + 
    geom_tile() + 
    scale_fill_distiller(
        palette = "Reds", 
        direction = 1) +
    labs(x = "q", y = "ICC")
```

```{r}
res_icc_q
```


```{r}
ggplot(as.data.frame(res_icc_q), 
       aes(x = as.character(q), 
           y = as.character(icc), 
           fill = points_mu)) + 
    geom_tile() + 
    scale_fill_distiller(
        palette = "Reds", 
        direction = 1) +
    labs(x = "q", y = "ICC")
```





```{r, fig.height=7, fig.width=7}
### separate group id and values
dat_id  = dat_sim %>% select(Group) 
dat_val = dat_sim %>% select(-Group)

### transpose the matrix and arrange
dat_id  = as.data.frame(dat_id)
dat_val = as.data.frame(t(dat_val))
colnames(dat_val) = rownames(dat_id)

### heatmap and labels
# add heatmap
setHook("grid.newpage", function(){
    pushViewport(
        viewport(
            x=1, y=1, width=0.9, height=0.9, 
            name="vp", just=c("right","top")))}, 
    action="prepend")

pheatmap(dat_val,
    annotation = dat_id,
    show_rownames = FALSE, 
    show_colnames = FALSE)

# add x, y labels
setHook("grid.newpage", NULL, "replace")
grid.text("Observations", y=-0.02, gp = gpar(fontsize=16))
grid.text("Variables",    x=-0.07, 
          rot = 90, gp=gpar(fontsize=16))
```


```{r}
dat
```


```{r}
foo()
```



## Q12. PCA and Clustering


# Part II. Working with Data

## Q21. Exploring / Clearing Data

## Q22. PCA

## Q23. Clustering

### K-Mediods

### Hierarchical Clustering

#### Clustering Method

#### Standardizing Data

#### Distance Metric

#### Sampling

#### Comment
Comment on what you think the most important factors are in cluster stability